---
title: "SecularDeathSim"
author: "Shawn Garbett"
date: "April 3, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The Problem

DES supports drawing from any observed distribution via sample to create patients. Modeling via differential equations requires a function. Since current simulations are using the observed data for establishment of a medical home. To convert this into a secular mortality probability that is a function of time for the simulation takes some effort.

## The data

The raw distributions taken from the Vanderbilt data sets, which was used for the PREDICT algorithm are as follows:

```{r studypop, echo=FALSE}
study_pop <- read.csv("../main/Age_Sex.csv")
study_pop.m <- subset(study_pop, gender=='M')$age.mh
study_pop.f <- subset(study_pop, gender=='F')$age.mh
par(mfrow=c(1,2))
hist(study_pop.m, main="Male", xlab="Age at Medical Home Start")
hist(study_pop.f, main="Female", xlab="Age at Medical Home Start")

```

## Strategy

Each individual's expected survival from social security can be overlaid into an array and this be used as the distribution of rates. We begin by fitting a spline to the 2011 social security data representing percent chance of death.
```{r spline, echo=FALSE}
ss_death <- read.csv("../mortality/ss-death-2011.csv", header=TRUE)

f_percent_d    <- c(ss_death$f_death_prob)
f_d_spline <- splinefun(ss_death$age + 0.5, f_percent_d)
curve(f_d_spline, col='red', from=0, to=120, xlab="Age", ylab="Probability of Secular Death")
m_percent_d    <- c(ss_death$m_death_prob)
m_d_spline <- splinefun(ss_death$age + 0.5, m_percent_d)
curve(m_d_spline, col='blue', add=TRUE)

# Construct it!
bins <- rep(0, 120) # Bin for each year possible
N    <- rep(0, 120)
for(i in 1:length(study_pop$age.mh))
{
  spl <- NULL
  spl <- if(study_pop[i, 'gender'] == 'F') f_d_spline else m_d_spline
  
  age <- study_pop[i,'age.mh']
  j   <- 1
  while(age < 120)
  {
    bins[j] <- bins[j] + spl(age)
    N[j]    <- N[j] + 1
    age     <- age + 1.0
    j       <- j + 1
  }
  bins[j]   <- bins[j] + 1.0
  N[j]      <- N[j] + 1
  
}
bins <- bins / N
```

The observed Vanderbilt's population's predicted rate of death post establishment of medical home is thus found to be the following:

```{r final, echo=FALSE}
plot(bins, typ='l', xlab="time in simulation", ylab="Probability of Secular Death")

final <- data.frame(time=0:102, death_rate=bins[1:103])

write.csv(final, "vandy-secular-death.csv", row.names=FALSE)
```

